Automatically Splitting a Two-Stage Lambda Calculus

### 2016-01-08 paper/abstract.tex
changes to front

Staged programming languages assign a stage to each program expression
and evaluate each expression in its assigned stage. A common use of
staged languages is to describe programs where inputs arrive at
different times or rates. In this paper we present an algorithm for
statically {\em splitting} these mixed-staged programs into two
unstaged, but dependent, programs where the outputs of the first
program can be efficiently reused across multiple invocations of the
second. While previous algorithms for performing this transformation
(also called \emph{pass separation} and \emph{data specialization})
were limited to operate on simpler, imperative languages, we define a
splitting algorithm for an explicitly-two-stage, typed lambda
calculus \lang\ with a $\fut$ modality denoting computation at a later
stage, and a $\curr$ modality noting purely first-stage code.  Most
notably, the algorithm splits mixed-stage recursive and higher-order
functions.  We prove the dynamic correctness of our splitting
algorithm with respect to a partial-evaluation semantics, and
mechanize this proof in Twelf. We also implement the algorithm
in a prototype compiler, and demonstrate that the ability to split
programs in a language featuring recursion and higher-order features
enables non-trivial algorithmic transformations that
improve code efficiency by up to a linear factor and also
facilitates modular expression of staged programs.

>>>>

Staged programming languages assign a stage to each program expression
and evaluate each expression in its assigned stage. A common use of
staged languages is to describe programs where inputs arrive at
different times or rates. In this paper we present an algorithm for
statically {\em splitting} these mixed-staged programs into two
unstaged, but dependent, programs where the outputs of the first
program can be efficiently reused across multiple invocations of the
second. While previous algorithms for performing this transformation
(also called \emph{pass separation} and \emph{data specialization})
were limited to operate on simpler, imperative languages, we define a
splitting algorithm for an explicitly-two-stage, typed lambda
calculus \lang\ with a $\fut$ modality denoting computation at a later
stage, and a $\curr$ modality noting purely first-stage code.  Most
notably, the algorithm splits mixed-stage recursive and higher-order
functions.  We prove the dynamic correctness of our splitting
algorithm with respect to a partial-evaluation semantics, and
mechanize this proof in Twelf. We also implement the algorithm
in a prototype compiler, and demonstrate that the ability to split
programs in a language featuring recursion and higher-order features
enables non-trivial algorithmic transformations that
improve code efficiency and also facilitates modular 
expression of staged programs.

### 2016-01-08 paper/examples.tex
mostly formatting
quick edit pass based on reviews
changes to front

1`dot : vec * vec -> $int
fun dot ((3`gr{`1`xliteralone`3`}`1`,`3`gr{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}),(3`gr{`1`xliteraltwo`3`}`1`,`3`gr{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`})) = 
  next{ 2`prev{`1`hold `3`gr{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)` 1`}`
\end{lstlisting}
%
The body of \texttt{dot} is a \rmint{} term at world \bbtwo\ containing a
\rmint\ computation of \texttt{(x1*x2) + (y1*y2)} which is promoted from world
\bbonep\ to world \bbtwo. We would expect the first stage of the split program
to take the first two coordinates of each vector and perform that first-stage
computation; and the second stage to take the final coordinates and the result
of the first stage, then multiply and add.

@@

This is an instance where stage-splitting is more practical than partial
evaluation---if $l$ is large, partially evaluating \texttt{quickselect} requires
runtime generation of a huge amount of code simultaneously encoding the tree and
traversal algorithm. (Avoiding the code blowup by not expanding some calls to
\texttt{part}, would result in duplicating first-stage computations.)

Thus, \texttt{qs1} performs $\Theta(n \log n)$ work per call, whereas
\texttt{qs2} performs expected $\Theta(\log n)$ work.  This results in a net
speedup over standard quickselect if we perform $\Omega(n / \log n)$ queries on
the same list.

>>>>

1`dot : vec * vec -> $int
fun dot ((3`gr{`1`xliteralone`3`}`1`,`3`gr{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}),(3`gr{`1`xliteraltwo`3`}`1`,`3`gr{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`})) = 
  next{ 2`prev{`1`hold `3`gr{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)` 1`}`
\end{lstlisting}
%
The body of \texttt{dot} is an \rmint\ term at world \bbtwo\ containing an
\rmint\ computation of \texttt{(x1*x2) + (y1*y2)} which is promoted from world
\bbonep\ to world \bbtwo. We would expect the first stage of the split program
to take the first two coordinates of each vector and perform that first-stage
computation; and the second stage to take the final coordinates and the result
of the first stage, then multiply and add.

@@

This is an instance where stage-splitting is more practical than partial
evaluation; if $l$ is large, partially evaluating \texttt{quickselect} requires
runtime generation of a huge amount of code simultaneously encoding the tree and
traversal algorithm. (Avoiding the code blowup by not expanding some calls to
\texttt{part}, would result in duplicating first-stage computations.)

Thus, \texttt{qs1} performs $\Theta(n \log n)$ work per call, whereas
\texttt{qs2} performs expected $\Theta(\log n)$ work.  This results in a net
speedup over standard quickselect if we perform $\omega(\log n)$ queries on
the same list.

### 2016-01-08 paper/overview.tex
changes to front

We can attempt to improve on this algorithm by factoring out computations shared
between these calls to \texttt{qs}. In particular, the comparisons performed by
\texttt{part} depend only on \texttt{l}, although different calls to
\texttt{part} are made by each call to \texttt{qs}.

A first cut at a solution might be to first sort \texttt{l} into
a sorted list \texttt{s} and then perform lookups in \texttt{s}.
Unfortunately, lookups in \texttt{s} are also $\Theta(n)$ on average, since
\texttt{s} is a list, leading to no improvement in efficiency.

A more sophisticated solution is to construct a binary search
tree out of \texttt{l} and then simply look for the \texttt{k}th
leftmost element of that tree. To do this efficiently requires one more
innovation---augmenting the tree by storing at each node the size of its left
subtree. For a sufficiently balanced tree, 
which one would get with a random choice of pivot, this approach has 
expected runtime $\Theta(n\log{n} + m\log{n})$, 
a nearly linear-time improvement in $m$.

In both cases, we replace our simple solution with a two-part algorithm which
first processes the input into a lookup data structure, and then subsequently
performs fast lookups. Improving algorithms in this way is highly non-trivial,
as it requires reasoning about intricate algorithmic concerns, and implementing
more complex algorithms and data structures.

In this paper, we present a splitting algorithm which automatically transforms
staged programs into multi-part programs. In certain cases, this algorithm even
improves input algorithms in the way described above!

>>>>

We can attempt to improve on this algorithm by factoring out computations shared
between these calls to \texttt{qs}, 
namely the parts that depend only on \texttt{l}.
A first cut at a solution might be to first sort \texttt{l} into
a sorted list \texttt{s}, at cost $\Theta(n \log n)$, and then perform lookups in \texttt{s}.
Unfortunately, lookups in \texttt{s} are $\Theta(n)$ on average, 
since \texttt{s} is a list.
Thus, the total cost is $\Theta(n \log n + mn)$, which is no improvement.

A more sophisticated solution is to construct a binary search
tree out of \texttt{l}, again at cost $\Theta(n\log{n})$, and then simply look for the \texttt{k}th
leftmost element of that tree. 
Doing lookups efficiently, in $\Theta(\log{n})$ time, 
requires one more innovation---augmenting the tree 
by storing at each node the size of its left subtree. 
Thus in total this approach has expected runtime $\Theta(n\log{n} + m\log{n})$.

Is this method, wherein we precompute a data structure, better than the direct $\Theta(mn)$ method from above? 
The answer depends on the relationship between the number of lookups $m$ and the size of the list $n$.
If $m$ is constant, then the direct method is superior for sufficiently large $n$.
This is because the precomputed method does unnecessary work 
sorting parts of the list where there are no query points.
However, if the number of queries grows with the size of the list, 
specficially in $\omega(\log n)$,
then the precomputed method will be asymptotically faster.

Rewriting algorithms in this way---to 
precompute some intermediate results that depend on constant (or slowly-varying) inputs---is 
highly non-trivial,
as it requires implementing more complex data structures and algorithms.  
In this paper, we present a splitting algorithm which does much of this task automatically.
